Co-evolution of lexical meaning \& pragmatic use

### 2017-03-29 paper/paper-v3.tex
revision TB + RvR

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. To investigate how
  cultural evolution of language plays out under this picture of the semantics-pragmatics
  interface, we present a game theoretic model of the competition between types of language
  users, each endowed with a selection of lexical concepts and a particular pragmatic
  disposition to act on them. Our model traces two evolutionary forces and their interaction:
  (i) fitness-based pressure towards communicative efficiency and (ii) potential learning
  biases during the transfer of linguistic knowledge. We illustrate the model based on a case
  study on scalar implicatures. In this case study learning biases that favor simple semantic
  representations can foster the evolution of more sophisticated pragmatic reasoning types and
  so prevent the lexicalization of scalar implicatures.
\end{abstract}

@@

\section{A model of co-evolving lexical concepts and pragmatic behavior}
\label{sec:model}

@@

Moving beyond an abstract example, our goal is to apply the RMD to investigate the co-evolution
of lexical concepts and pragmatic behavior. To do so, we need to fix three things: (i) what the
relevant types are, (ii) how fitness derives from communicative success and (iii) how the
mutation matrix $Q$ is computed. These issues are addressed, one by one, in the following.

@@

Types are what evolution operates on. They define an agent's fitness, usually through a payoff
accrued in single interactions with other agents. Often types can be identified as the possible
acts in a game, e.g., either cooperating or defecting in a prisoner's dilemma. In other cases,
they may be thought of as general properties of an agent that influences her fitness, such as
being positive or negative in our previous example (whatever that means). For our present
purposes, types are identified more concretely by specific assumptions about their cognitive
make-up. Since we are interested in the evolutionary competition between different lexical
concepts and ways of using them in communication, a type is here defined as a pair consisting
of a lexicon and a pragmatic strategy of language use.

@@

It is particularly important to note that pragmatic agents of level 1 using $\Llack$ associate $\msome$
preferentially with $\ssome$, in contrast to their literal counterparts of level 0. This
association is not perfect, and usually less strong than what agents with a lexicalized upper
bound in $\Lbound$ can achieve, even without pragmatic reasoning. Higher order 
reasoning types leads to stronger associations of $\msome$ and $\ssome$ also for the
hearer. Still, the case study presented in Section~\ref{sec:si-case-study} will consider
speaker and hearer behavior at levels 0 and 1, as the latter are the simplest pragmatic
reasoning types which show a tendency to communicatively attuned pragmatic enrichment. Using
only level 1 reasoning types is therefore a conservative choice that works against the
fitness-based selection of pragmatic language use for a notion of fitness based on
communicative success, which is introduced next.

@@

Following \citet{griffiths+kalish:2007} we model language acquisition as a process of Bayesian
inference in which learners combine the likelihood of a type producing the witnessed learning
input with prior inductive biases. Experimental and mathematical results on iterated learning
suggest that the outcome of this process reflects learners' inductive biases
\citep[e.g.,][]{kirby+etal:2014}. In a Bayesian setting these biases can be codified in a prior
$P \in \Delta(T)$, which reflects the amount of data a learner requires to faithfully acquire
the language of the teacher \citep[cf.][450]{griffiths+kalish:2007}. The extent of the prior's
influence has been shown to heavily depend on the learning strategy assumed to underlie the
inference process. On the one hand, early simulation results suggested that weak biases could
be magnified by exposing learners to only small data samples
\citep[e.g. in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
characterization showed that iterated learning converges to the prior in the limit, i.e., that
the resulting distribution over languages corresponds to the learners' prior distribution and
is not influenced by the amount of input given to them. This difference in predictions can be
traced back to differences in the selection of hypotheses from the posterior. Griffith \&
Kalish's convergence to the prior holds for learners that sample from the posterior. More
deterministic strategies such as the adoption of the type with the highest posterior
probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of
both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we
use a parameter $l\ge1$ to modulate between posterior sampling and the MAP strategy. When
$l = 1$ learners sample from the posterior. The learners' propensity to maximize the posterior
grows as $l$ increases.

@@

The model of the previous section formalizes the evolutionary competition between different
pairs of lexical concepts and ways of using them. This section looks at a case study on scalar
implicatures. It engages in a following formal thought experiment to address the question: if a
population of language users could freely combine different lexica with different pragmatic
strategies, what are conditions under which the majority view of scalar implicatures could have
evolved?

@@

To fill the model from Section~\ref{sec:model} with life, we need to specify the sets of
states, messages and lexica (Section~\ref{sec:stat-mess-conc}). Additionally, we want to
explore the effects of a learning bias in favor of simple lexical concepts. One way of
motivating and formalizing such a bias is introduced below in
Section~\ref{sec:an-induct-learn}.

\subsubsection{States, messages, concepts and lexica} 
\label{sec:stat-mess-conc}

Consider a state space with three states $\States = \set{\snone, \ssome, \sall}$ and think of
it as a partition of possible worlds into cells where none, some or all of the $A$s are $B$s,
for some arbitrary fixed predicates $A$ and $B$. Eight concepts can be distinguished based on
their truth or falsity in three world states, six of which are not contradictory or
tautological (see Table~\ref{tab:concepts} below).

A lexicon $L$ is a mapping $\Messgs \rightarrow C$ from messages to concepts. With three
messages there are $6^3 = 216$ possible lexica. Some assign the same concept to more than one
message and others lexicalize the same concepts but associate them with different messages. Out
of these possible lexica, three kinds are of particular relevance. First, lexica that assign
the same concept to more than one message. Such lexica lack in expressivity but may be favored
by particular learning biases nonetheless (see below). Second, lexica that conventionalize
upper-bounds to realize a one-to-one mapping of messages to states. Finally, lexically that do
not lexicalize an upper bound but allow it to be conveyed pragmatically due to the presence of
a stronger lexical item. There are six lexica of the second kind and six of the third. The
following three lexica exemplify each kind:
\begin{align*}
  \begin{blockarray}{lccc}
    & & \underline{\Lall} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
     \mygray{\snone}  & 0 & 0 & 0\\
     \mygray{\ssome}  & 0 & 0 & 0\\
    \mygray{\sall}   & 1 & 1 & 1\\
    \end{block}
% \\ & & P(\Lall)\rlap{ $\propto 9$} & \\
  \end{blockarray} & &
 \begin{blockarray}{lccc}
    & & \underline{\Lbound} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 0 & 1\\
    \end{block}
%  \\& & P(\Lbound)\rlap{ $\propto 96$} & \\
  \end{blockarray} & &
  \begin{blockarray}{lccc}
    & & \underline{\Llack} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 1 & 1\\
    \end{block}
%   \\& & P(\Llack)\rlap{ $\propto 48$} & \\
  \end{blockarray}
\end{align*}
Lexicon $\Lall$ is clearly bad for communication: all message and interpretation choices will
be equally likely for all types; no information about the observed world state will be conveyed
by its users. In contrast, users of $\Lbound$ can communicate world states perfectly, no matter
whether they are literal or pragmatic users. Users of $\Llack$ can also communicate information
about the actual world state but need pragmatic language use to approximate a one-to-one
mapping between message use and states (see Section~\ref{sec:languages+use}).

Recall that types are a combination of a lexicon and a manner of language use. We analyze the
model's predictions in populations of types with one of the two behaviors introduced earlier;
literal or pragmatic. The former correspond to level $0$ reasoners and the latter to level
$1$. Accordingly, we consider a total of $432$ types. Six are variants of pragmatic language
users with $\Llack$-like lexica. We refer to these as \emph{target types}, because they
represent lexica and language use that conform to the majority view of scalar
implicatures. Twelve types are either literal or pragmatic speakers with lexica of the
$\Lbound$ kind. We refer to these as \emph{competitor types}, because they are expected to be
the target types' main contenders in evolutionary competition.

@@

There is a growing effort to develop empirically testable representational languages that allow
for the measure of semantic complexity. For instance, so-called {\em languages of thought}
(LOTs) have been put to test in various rational probabilistic models that show encouraging
results (see, e.g., \citealt{katz+etal:2008}, \citeauthor{piantadosi+etal:underreview} under review; 2012, and \citealt{piantadosi+jacobs:2016} for recent discussion). At its
core, a LOT defines a set of operations and composition rules from which concepts can be
derived. As a first approximation and for the sake of concreteness, we follow this approach to
motivate and formalize a preference of learners for simpler semantic representations
(\citealt{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015}, \citeauthor{piantadosi+etal:underreview} under review). 
In a weighed generative LOT a concept's complexity is a function of its derivation cost.

@@

Our toy grammar of concepts is given in Table~\ref{tab:grammar}. This grammar uses basic
set-theoretic operations to form expressions which can be evaluated as true or false in states
$\snone$, $\ssome$, or $\sall$ from above. Applications of generative rules have a cost attached
to them. Here we simply assume that the formation of Boolean combinations of concepts incurs 2
cost units, while all other rule applications incur only 1 cost unit. Table~\ref{tab:concepts}
lists all six concepts relevant here, their truth conditions and the simplest formula that
expresses this concept from the grammar in Table~\ref{tab:grammar}.

\begin{table}
  \centering
\begin{center}
  \begin{tabular}{lccclc}
    \toprule
    intuitive name
    & \snone
    & \ssome
    & \sall
    & least complex formula
    & complexity
    \\ \midrule
    ``all''
    & 0
    & 0
    & 1
    & $A \subseteq B$
    & $3$
    \\
    ``some but not all''
    & 0
    & 1
    & 0
    & $A \cap B \neq \emptyset \wedge A \neq \emptyset$
    & $8$
    \\    
    ``some''
    & 0
    & 1
    & 1
    & $A \cap B \neq \emptyset$
    & $4$
    \\
    ``none''
    & 1
    & 0
    & 0
    & $A \cap B = \emptyset$
    & $4$
    \\
    ``none or all''
    & 1
    & 0
    & 1
    & $\neg(A \cap B \neq \emptyset \wedge A \neq \emptyset)$
    & $10$
    \\
    ``not all''
    & 1
    & 1
    & 0
    & $\neg (A \subseteq B)$
    & $5$
    \\
    \bottomrule
  \end{tabular}
\end{center}
\caption{Available concepts and their minimal derivation cost}
\label{tab:concepts}
\end{table}

A complexity measure for lexical concepts from Table~\ref{tab:concepts} is used to define a
learning bias that favors simpler concepts over more complex ones. The prior probability of a
type is just the prior probability of its lexicon. The prior of a lexicon is a function of the
complexity of the lexical representations in its image set. Lexica with simpler concepts accordingly have a
higher prior. One simple way of defining such priors over lexica (and thereby types) is:
\begin{align*}
  P(L)  & \propto \prod_{c \in Im(L)} P(c)   \ \text{, with} & 
  P(c) & \propto \max_{c'}Compl(c') - Compl(c) + 1\,,
\end{align*}
where $Compl(c)$ is the complexity of the minimal derivation cost of $c$ according to the
LOT-grammar (see Table~\ref{tab:concepts}). Applied to our space of lexica, this construal assigns the highest probability 
to a lexicon of type $\Lall$, which only uses the simplest concept ``all'' for
all messages. Lexica of type $\Llack$ are less likely, but more likely than $\Lbound$.

@@

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
\caption{Stacked proportion of target types and majority types, when other
  than target type, in $5$ independent populations after $50$ generations under only a
  pressure for communicative success. The right-most plots zoom in on only the proportion of target types.}
\label{fig:only-R}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
\caption{Stacked proportion of target types and majority types, when other than target type, in $5$ independent populations after $50$ generations under only a pressure for learnability ($\lambda = 20, k = 5$). The types' prior is shown in
  the right-most plot. The top-most groupings correspond to types with lexica of the kind of $\Llack$ (green), $\Lbound$ (magenta), and $\Lall$ (gray).}
\label{fig:only-M}
\end{figure}

@@

Pressure for communicative success or learnability are not sufficient on their own to have a
single target type dominate the population. When pressured for communicative success, the slight
communicative advantage of $\Lbound$ users leads to the its prevalence. When pressured for
learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
complex alternatives such as $\Lbound$. However, learnability alone does not foment the
propagation of a single target type across the population.

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig3-r+m}
\caption{Stacked proportion of target types and majority types, when other than target type, in $5$ independent populations after $50$ generations under both pressures ($k = 5$).}
\label{fig:rmd}
\end{figure}

@@

The main result of our case study is that types that correspond to the majority view of scalar
implicatures (scalar readings are non-lexicalized pragmatic enrichments) can come to dominate a
population. This can happen under the assumption that simpler semantic representations are more
likely to be learned (cf. \citealt{chater+vitanyi:2003}). Pragmatic language use can be
recruited indirectly by a preference for simpler lexical concepts. Under this view, semantics
and pragmatics play a synergic role: pragmatic use allows maintenance of simpler concepts;
pressure towards representational simplicity indirectly promotes pragmatic over literal
language use. As a consequence, iterated transmission and use of language lead to a
regularization that may explain the lack of lexicalization of systematic pragmatic enrichments.

While the results of this case study are interesting, they also raise a number of critical
issues. First of all, while many favorable parameter settings exist which lead to a prevalence
of target types, other types are usually represented in non-negligible proportions. This may
just be a technical quirk of the mutator step. But there is a related issue of empirical
importance. Several experimental studies on scalar implicatures suggest that participants can
be classified as either semantic or pragmatic users of, in particular, \emph{some}
\citep[e.g.][]{BottNoveck2004:Some-Utterances,NieuwlandDitman2010:On-the-incremen,DegenTanenhaus2012:Processing-Scal}. The
former consistently accept \emph{some} where \emph{all} would be true as well, the latter do
not. Interestingly, in our simulations when a target type is the majority type an inflated
proportion of the population uses compatible lexica with a lexicalized upper bound. Particularly in those parameter
settings where the prior influences the outcome less. In other words, we do find a tendency toward a similar co-existence
of semantic and pragmatic types. Whether this analogy has any further explanatory value is an interesting path for future
exploration. \tb{While it is indeed true that $\Lbound$'s proportion is inflated relative to other types in the hypothesis space, it is never the second most prolific type (which is what the above hopefully is not taken to suggest, otherwise it should be changed). As is to be expected, $\Lbound$ fares best when $l$ is low and $\lambda$ is high. Even then, it usually ranks around $5$-th with a mean proportion of its kind of about $0.01 - 0.05$. As $l$ increases, it gets pushed back. Second or third place are usually instead literal $\Llack$ or types that lexicalize ``all'', ``not all'' and ``none''. In short; yes, there is a inflated proportion relative to the whole population, but other types are usually more prolific. This is very likely due to the strong prior disadvantage of $\Llack$. In light of this I'm uncertain whether we want to keep this paragraph as is. I currently tend towards `yes' but let me know if you think otherwise.}

Another important issue that is not addressed in the model are potential costs associated with
pragmatic reasoning. Here, we simply assumed that literal and pragmatic reasoning strategies
exist from the start and are equally costly to apply. In contrast, empirical results suggest
that the computation of a scalar implicature may involve additional cognitive effort
\citep[e.g.][]{BrehenyKatsos2006:Are-Generalised,deNeys+schaeken:2007,huang+snedeker:2009,Jr.Bailey2013:Possibly-all-of}. Extensions
of the model presented here to include processing costs for pragmatic language use would be
interesting future work. It seems plausible that effects of reasoning cost may trade off with
the frequency with which a given scalar expression is used. It may be that frequently drawn
scalar implicatures lexicalize to avoid cost, whereas infrequent ones are derived on-line to
avoid more complex lexical concepts during acquisition. Such a prediction would lend itself to
empirical testing in line with a recent interest in differences between various scalar
implicature triggers \citep{Tielvan-TielMiltenburgvan-Miltenburg2014:Scalar-Diversit}.

Our case study could be criticized as follows: all it shows is that scalar implicatures do not
lexicalize because upper bounds are dispreferred concepts. This criticism would be too
superficial and highly unjust. Dispreferred lexical concepts can thrive under evolutionary
selection. Lexicalized upper-bounds can dominate a population because they may boost
communicative success. But they do not have to. Moreover, even without selective pressure
for communicative success, it is not the case that necessarily the types that are most
likely \emph{a priori} will dominate. The dynamics of iterated learning are not that
trivial. Iterated learning does not necessarily promote the \emph{a priori} more likely type,
but tends to promote a type $t$ based on a gradient of how many other types might likely mutate
into $t$, so to speak. Taken together, without an explicit model of the interaction between
pressure for success and learnability, it is far from trivial to judge whether or when
preferred or dispreferred concepts can be adaptive. This is why a major contribution of this
paper is the arrangement of many different ingredients into a joined model of the co-evolution
of lexical meaning and pragmatic use.

>>>>

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. To investigate how
  cultural evolution of language plays out under this picture of the semantics-pragmatics
  interface, we present a game theoretic model of the competition between types of language
  users, each endowed with a selection of lexical representations and a particular pragmatic
  disposition to act on them. Our model traces two evolutionary forces and their interaction:
  (i) fitness-based pressure towards communicative efficiency and (ii) potential learning
  biases during the transfer of linguistic knowledge. We illustrate the model based on a case
  study on scalar implicatures. In this case study learning biases that favor simple semantic
  representations can foster the evolution of more sophisticated pragmatic reasoning types and
  so prevent the lexicalization of scalar implicatures.
\end{abstract}

@@

\section{A model of co-evolving lexical representations and pragmatic behavior}
\label{sec:model}

@@

Moving beyond an abstract example, our goal is to apply the RMD to investigate the co-evolution
of lexical representations and pragmatic behavior. To do so, we need to fix three things: (i) what the
relevant types are, (ii) how fitness derives from communicative success and (iii) how the
mutation matrix $Q$ is computed. These issues are addressed, one by one, in the following.

@@

Types are what evolution operates on. They define an agent's fitness, usually through a payoff
accrued in single interactions with other agents. Often types can be identified as the possible
acts in a game, e.g., either cooperating or defecting in a prisoner's dilemma. In other cases,
they may be thought of as general properties of an agent that influences her fitness, such as
being positive or negative in our previous example (whatever that means). For our present
purposes, types are identified more concretely by specific assumptions about their cognitive
make-up. Since we are interested in the evolutionary competition between different lexical
representations and ways of using them in communication, a type is here defined as a pair consisting
of a lexicon and a pragmatic strategy of language use.

@@

It is particularly important to note that pragmatic agents of level 1 using $\Llack$ associate $\msome$
preferentially with $\ssome$, in contrast to their literal counterparts of level 0. This
association is not perfect, and usually less strong than what agents with a lexicalized upper
bound in $\Lbound$ can achieve, even without pragmatic reasoning. Higher order 
reasoning beyond level $1$ leads to stronger associations of $\msome$ and $\ssome$ also for the
hearer. Still, the case study presented in Section~\ref{sec:si-case-study} will consider
speaker and hearer behavior at levels 0 and 1, as the latter are the simplest pragmatic
reasoning types which show a tendency to communicatively attuned pragmatic enrichment. Using
only level 1 reasoning types and possibly small $\lambda$ is therefore a conservative choice that works against the
fitness-based selection of pragmatic language use for a notion of fitness based on
communicative success, which is introduced next.

@@

Following \citet{griffiths+kalish:2007} we model language acquisition as a process of Bayesian
inference in which learners combine the likelihood of a type producing the witnessed learning
input with prior inductive biases. In a Bayesian setting these biases can be codified in a prior
$P \in \Delta(T)$, which reflects the amount of data a learner requires to faithfully acquire
the language of the teacher \citep[cf.][450]{griffiths+kalish:2007}. The extent of the prior's
influence has been shown to heavily depend on the learning strategy assumed to underlie the
inference process. On the one hand, early simulation results suggested that weak biases could
be magnified by exposing learners to only small data samples
\citep[e.g. in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
characterization showed that iterated learning converges to the prior in the limit, i.e., that
the resulting distribution over languages corresponds to the learners' prior distribution and
is not influenced by the amount of input given to them. This difference in predictions can be
traced back to differences in the selection of hypotheses from the posterior. Griffith \&
Kalish's convergence to the prior holds for learners that sample from the posterior. More
deterministic strategies such as the adoption of the type with the highest posterior
probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of
both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we
use a parameter $l\ge1$ to modulate between posterior sampling and the MAP strategy. When
$l = 1$ learners sample from the posterior. The learners' propensity to maximize the posterior
grows as $l$ increases.

@@

The model of the previous section formalizes the evolutionary competition between different
pairs of lexical representations and ways of using them. This section looks at a case study on scalar
implicatures. It engages in a following formal thought experiment to address the question: if a
population of language users could freely combine different lexica with different pragmatic
strategies, what are conditions under which the majority view of scalar implicatures could have
evolved?

@@

To fill the model from Section~\ref{sec:model} with life, we need to specify the sets of
states, messages and lexica (Section~\ref{sec:stat-mess-conc}). Additionally, we want to
explore the effects of a learning bias in favor of simple lexical representations. One way of
motivating and formalizing such a bias is introduced below in
Section~\ref{sec:an-induct-learn}.

\subsubsection{States, messages, lexical representations, and lexica} 
\label{sec:stat-mess-conc}

Consider a state space with three states $\States = \set{\snone, \ssome, \sall}$ and think of
it as a partition of possible worlds into cells where none, some or all of the $A$s are $B$s,
for some arbitrary fixed predicates $A$ and $B$. Eight lexical representations can be distinguished based on
their truth or falsity in three world states, six of which are not contradictory or
tautological (see Table~\ref{tab:concepts} below).

A lexicon $L$ is a mapping $\Messgs \rightarrow R$ from messages to representations. With three
messages there are $6^3 = 216$ possible lexica. Some assign the same representations to more than one
message and others lexicalize the same representations but associate them with different messages. Out
of these possible lexica, three kinds are of particular relevance. First, lexica that assign
the same lexical representations to more than one message. Such lexica lack in expressivity but may be favored
by particular learning biases nonetheless (see below). Second, lexica that conventionalize
upper-bounds to realize a one-to-one mapping of messages to states. Finally, lexically that do
not lexicalize an upper bound but allow it to be conveyed pragmatically due to the presence of
a stronger lexical item. There are six lexica of the second kind and six of the third. 
The following three lexica exemplify each kind:
\begin{align*}
  \begin{blockarray}{lccc}
    & & \underline{\Lall} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
     \mygray{\snone}  & 0 & 0 & 0\\
     \mygray{\ssome}  & 0 & 0 & 0\\
    \mygray{\sall}   & 1 & 1 & 1\\
    \end{block}
% \\ & & P(\Lall)\rlap{ $\propto 9$} & \\
  \end{blockarray} & &
 \begin{blockarray}{lccc}
    & & \underline{\Lbound} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 0 & 1\\
    \end{block}
%  \\& & P(\Lbound)\rlap{ $\propto 96$} & \\
  \end{blockarray} & &
  \begin{blockarray}{lccc}
    & & \underline{\Llack} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 1 & 1\\
    \end{block}
%   \\& & P(\Llack)\rlap{ $\propto 48$} & \\
  \end{blockarray}
\end{align*}
Lexicon $\Lall$ is clearly bad for communication: all message and interpretation choices will
be equally likely for all types; no information about the observed world state will be conveyed
by its users. In contrast, users of $\Lbound$ can communicate world states perfectly, no matter
whether they are literal or pragmatic users. Users of $\Llack$ can also communicate information
about the actual world state but need pragmatic language use to approximate a one-to-one
mapping between message use and states (see Section~\ref{sec:languages+use}).

Recall that types are a combination of a lexicon and a manner of language use. We analyze the
model's predictions in populations of types with one of the two behaviors introduced earlier;
literal or pragmatic. The former correspond to level $0$ reasoners and the latter to level
$1$. Accordingly, we consider a total of $432$ types. Six are variants of pragmatic language
users with $\Llack$-like lexica. We refer to these as \emph{target types}, because they
represent lexica and language use that conform to the majority view of scalar
implicatures. Twelve types are either literal or pragmatic speakers with lexica of the
$\Lbound$ kind. We refer to these as \emph{competitor types}, because they are expected to be
the target types' main contenders in evolutionary competition. Finally, note that while variants of a kind
lexicalize the same representations and use them in the same way when paired with identical linguistic behavior,
they map different states to different messages. More informally, they speak different languages
that lexicalize the same concepts. Consequently, more often than not types of a kind fail to communicate with each other.

@@

There is a growing effort to develop empirically testable representational languages that allow
for the measure of semantic complexity. For instance, so-called {\em languages of thought}
(LOTs) have been put to test in various rational probabilistic models that show encouraging
results (see, e.g., \citealt{katz+etal:2008}, \citeauthor{piantadosi+etal:underreview} under review; 2012, and \citealt{piantadosi+jacobs:2016} for recent discussion). At its
core, a LOT defines a set of operations and composition rules from which lexical representations can be
derived. As a first approximation and for the sake of concreteness, we follow this approach to
motivate and formalize a preference of learners for simpler semantic representations
(\citealt{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015}, \citeauthor{piantadosi+etal:underreview} under review). 
In a weighed generative LOT a representation's complexity is a function of its derivation cost.

@@

Our toy grammar of lexical representations is given in Table~\ref{tab:grammar}. This grammar uses basic
set-theoretic operations to form expressions which can be evaluated as true or false in states
$\snone$, $\ssome$, or $\sall$ from above. Applications of generative rules have a cost attached
to them. Here we simply assume that the formation of Boolean combinations of representations incurs 2
cost units, while all other rule applications incur only 1 cost unit. Table~\ref{tab:concepts}
lists all six lexical representations relevant here, their truth conditions and the simplest formula that
expresses this representation from the grammar in Table~\ref{tab:grammar}.

\begin{table}
  \centering
\begin{center}
  \begin{tabular}{lccclc}
    \toprule
    intuitive name
    & \snone
    & \ssome
    & \sall
    & least complex formula
    & complexity
    \\ \midrule
    ``all''
    & 0
    & 0
    & 1
    & $A \subseteq B$
    & $3$
    \\
    ``some but not all''
    & 0
    & 1
    & 0
    & $A \cap B \neq \emptyset \wedge A \neq \emptyset$
    & $8$
    \\    
    ``some''
    & 0
    & 1
    & 1
    & $A \cap B \neq \emptyset$
    & $4$
    \\
    ``none''
    & 1
    & 0
    & 0
    & $A \cap B = \emptyset$
    & $4$
    \\
    ``none or all''
    & 1
    & 0
    & 1
    & $\neg(A \cap B \neq \emptyset \wedge A \neq \emptyset)$
    & $10$
    \\
    ``not all''
    & 1
    & 1
    & 0
    & $\neg (A \subseteq B)$
    & $5$
    \\
    \bottomrule
  \end{tabular}
\end{center}
\caption{Available lexical representations and their minimal derivation cost}
\label{tab:concepts}
\end{table}

A complexity measure for lexical representations from Table~\ref{tab:concepts} is used to define a
learning bias that favors simpler representations over more complex ones. The prior probability of a
type is just the prior probability of its lexicon. The prior of a lexicon is a function of the
complexity of the lexical representations in its image set. Lexica with simpler representations accordingly have a
higher prior. One simple way of defining such priors over lexica (and thereby types) is:
\begin{align*}
  P(L)  & \propto \prod_{c \in Im(L)} P(c)   \ \text{, with} & 
  P(c) & \propto \max_{c'}Compl(c') - Compl(c) + 1\,,
\end{align*}
where $Compl(c)$ is the complexity of the minimal derivation cost of $c$ according to the
LOT-grammar (see Table~\ref{tab:concepts}). Applied to our space of lexica, this construal assigns the highest probability 
to a lexicon of type $\Lall$, which only uses the simplest lexical representation ``all'' for
all messages. Lexica of type $\Llack$ are less likely, but more likely than $\Lbound$.

@@

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
\caption{Stacked proportion of target types, in shades of green, and majority types, when other
  than target type, in $5$ independent populations after $50$ generations under only a
  pressure for communicative success. The right-most plots zoom in on only the proportion of target types.}
\label{fig:only-R}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
\caption{Stacked proportion of target types, in shades of green, and majority types, when other than target type, in $5$ independent populations after $50$ generations under only a pressure for learnability ($\lambda = 20, k = 5$). The types' prior is shown in
  the right-most plot. The top-most groupings correspond to types, literal and pragmatic, with lexica of the kind of $\Llack$, $\Lbound$, and $\Lall$.}
\label{fig:only-M}
\end{figure}

@@

Pressure for communicative success or learnability are not sufficient on their own to have a
single target type dominate the population. When pressured for communicative success, the slight
communicative advantage of $\Lbound$ users leads to the their prevalence. When pressured for
learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
complex alternatives such as $\Lbound$. However, learnability alone does not foment the
propagation of a single target type across the population.

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig3-r+m}
\caption{Stacked proportion of target types, in shades of green, and majority types, when other than target type, in $5$ independent populations after $50$ generations under both pressures ($k = 5$).}
\label{fig:rmd}
\end{figure}

@@

The main result of our case study is that types that correspond to the majority view of scalar
implicatures (scalar readings are non-lexicalized pragmatic enrichments) can come to dominate a
population. This can happen under the assumption that simpler semantic representations are more
likely to be learned (cf. \citealt{chater+vitanyi:2003}). Pragmatic language use can be
recruited indirectly by a preference for simpler lexical representations. Under this view, semantics
and pragmatics play a synergic role: pragmatic use allows maintenance of simpler representations;
pressure towards representational simplicity indirectly promotes pragmatic over literal
language use. As a consequence, iterated transmission and use of language lead to a
regularization that may explain the lack of lexicalization of systematic pragmatic enrichments.

While the results of this case study are interesting, they also raise a number of critical
issues. First of all, while many favorable parameter settings exist which lead to a prevalence
of target types, other types are usually represented in non-negligible proportions. This may
just be a technical quirk of the mutator step. But there is a related issue of empirical
importance. Several experimental studies on scalar implicatures suggest that participants can
be classified as either semantic or pragmatic users of, in particular, \emph{some}
\citep[e.g.][]{BottNoveck2004:Some-Utterances,NieuwlandDitman2010:On-the-incremen,DegenTanenhaus2012:Processing-Scal}. The
former consistently accept \emph{some} where \emph{all} would be true as well, the latter do
not. Interestingly, in our simulations when a target type is the majority type an inflated
proportion of the population uses compatible lexica with a lexicalized upper bound. Particularly in those parameter
settings where the prior influences the outcome less. In other words, we do find a tendency toward a similar co-existence
of semantic and pragmatic types. Whether this analogy has any further explanatory value is an interesting path for future
exploration. \tb{While it is indeed true that $\Lbound$'s proportion is inflated relative to other types in the hypothesis space, it is never the second most prolific type (which is what the above hopefully is not taken to suggest, otherwise it should be changed). As is to be expected, $\Lbound$ fares best when $l$ is low and $\lambda$ is high. Even then, it usually ranks around $5$-th with a mean proportion of its kind of about $0.01 - 0.05$. As $l$ increases, it gets pushed back. Second or third place are usually instead literal $\Llack$ or types that lexicalize ``all'', ``not all'' and ``none''. In short; yes, there is a inflated proportion relative to the whole population, but other types are usually more prolific. This is very likely due to the strong prior disadvantage of $\Llack$. In light of this I'm uncertain whether we want to keep this paragraph as is. I currently tend towards `yes' but let me know if you think otherwise. Robert would leave it in as well.}

Another important issue that is not addressed in the model are potential costs associated with
pragmatic reasoning. Here, we simply assumed that literal and pragmatic reasoning strategies
exist from the start and are equally costly to apply. In contrast, empirical results suggest
that the computation of a scalar implicature may involve additional cognitive effort
\citep[e.g.][]{BrehenyKatsos2006:Are-Generalised,deNeys+schaeken:2007,huang+snedeker:2009,Jr.Bailey2013:Possibly-all-of}. Extensions
of the model presented here to include processing costs for pragmatic language use would be
interesting future work. It seems plausible that effects of reasoning cost may trade off with
the frequency with which a given scalar expression is used. It may be that frequently drawn
scalar implicatures lexicalize to avoid cost, whereas infrequent ones are derived on-line to
avoid more complex lexical representations during acquisition. Such a prediction would lend itself to
empirical testing in line with a recent interest in differences between various scalar
implicature triggers \citep{Tielvan-TielMiltenburgvan-Miltenburg2014:Scalar-Diversit}.

Our case study could be criticized as follows: all it shows is that scalar implicatures do not
lexicalize because upper bounds are dispreferred lexical representations. This criticism would be too
superficial and highly unjust. Dispreferred lexical representations can thrive under evolutionary
selection. Lexicalized upper-bounds can dominate a population because they may boost
communicative success. But they do not have to. Moreover, even without selective pressure
for communicative success, it is not the case that necessarily the types that are most
likely \emph{a priori} will dominate. The dynamics of iterated learning are not that
trivial. Iterated learning does not necessarily promote the \emph{a priori} more likely type,
but tends to promote a type $t$ based on a gradient of how many other types might likely mutate
into $t$, so to speak. Taken together, without an explicit model of the interaction between
pressure for success and learnability, it is far from trivial to judge whether or when
preferred or dispreferred representations can be adaptive. This is why a major contribution of this
paper is the arrangement of many different ingredients into a joined model of the co-evolution
of lexical meaning and pragmatic use.
